some title

### 2018-01-07 2018/images/corr_def.tex
corr and r^2

\usepackage{tikz-3dplot} %requires 3dplot.sty to be in same directory, or in your LaTeX installation
\usepackage{tkz-euclide}
\usetkzobj{all}
\usepackage[active,tightpage]{preview}  %generates a tightly fitting border around the work
\PreviewEnvironment{tikzpicture}
\setlength\PreviewBorder{2mm}

\begin{document}

%Angle Definitions
%-----------------

%start tikz picture, and use the tdplot_main_coords style to implement the display
%coordinate transformation provided by 3dplot
\tdplotsetmaincoords{40}{120}

\begin{tikzpicture}[tdplot_main_coords]

\coordinate (O) at (0,0,0);
\coordinate (xc) at (2.1,-2.1);
\coordinate (xcp) at (0.84,-0.84);
\coordinate (yc) at (-0.9,3);
\coordinate (un) at (0,0,6);
\coordinate (unt) at (0,0,2.9);

\tdplotsetcoord{S}{10}{35}{0}  % икс
\tdplotsetcoord{R}{4}{35}{0} % разложение
\tdplotsetcoord{P}{7}{30}{90} % игрек
\tdplotsetcoord{T}{6.5}{20}{0} % икс с тильдой

\draw[->] (O) -- (un) node[anchor=east]{$\mathbf{1}$};
\draw[->] (O) -- (S) node[anchor=south east]{$x$};
\draw[->, gray] (O) -- (P) node[anchor=north west, gray]{$y$};
\draw[->] (O) -- (T) node[anchor=south east]{$\tilde x$};

\draw (0,0.5) ellipse (4cm and 1cm) node[anchor=west, right=2.7cm]{\tiny$Lin^{\perp}(\mathbf{1})$};

\draw[->, gray] (O) -- (yc) node[anchor=north east]{\small $y^c$};
\draw[->] (O) -- (xc) node[anchor=north east]{\small $x^c$};
\draw[dashed, gray] (P) -- (yc);
\draw[dashed] (S) -- (xc);

\draw[->, dashed, blue] (unt) -- (T);
\draw[->, dashed, red] (R) -- (T);
\draw[->, blue] (O) -- (R) node[anchor=north east]{\footnotesize$\beta x$};
\draw[->, red] (O) -- (unt) node[anchor=west]{\footnotesize$\alpha \mathbf{1}$};

\draw[dashed, blue] (R) -- (xcp);
\draw[->,dashed,blue] (O) -- (xcp) node[anchor=north]{\small$\tilde x^c$};
\tkzMarkRightAngle[size=0.15,color=blue](R,xcp,O);
\tkzDrawPoint[color=red,fill=red,size=1.8](O);

\tkzMarkRightAngle[size=0.15,color=gray](P,yc,O);
\tkzMarkRightAngle[size=0.15](S,xc,O);
\tdplotdrawarc{(O)}{0.15}{-45}{110}{anchor=north}{\tiny$\varphi$}

>>>>

\usepackage{tikz-3dplot} %requires 3dplot.sty to be in same directory, or in your LaTeX installation
\usepackage{tkz-euclide}
\usetkzobj{all}
\usetikzlibrary{calc}
\usepackage[active,tightpage]{preview}  %generates a tightly fitting border around the work
\PreviewEnvironment{tikzpicture}
\setlength\PreviewBorder{2mm}

\begin{document}
\begin{tikzpicture}

\tkzDefPoint(0,0){O}
\tkzDefPoint(4,0){X}
\tkzDefPoint(2,3){Y}

\draw[->] (O) -- (X) node[anchor=north east]{$x$};
\draw[->] (O) -- (Y) node[anchor=east]{$y$};

\tkzMarkAngle[size=0.4](X,O,Y)
\tkzLabelAngle[pos=0.6](X,O,Y){$\varphi$}

### 2018-01-07 2018/paper.tex
corr and r^2

\subsection{Correlation}

\marginnote{The most widespread definition of correlation is
\[
\Corr(X,Y)  = \frac{\Cov(X, Y)}{\sqrt{\Var(X)\Var(Y)}}
\]}

The most crucial part in defining correlation geometrically is definig the dot product as it enables to compute the length of a vecotr:
\[
|\vec{a}| = \sqrt{\langle \vec{a},  \vec{a}\rangle}
\]
and the angle between any two vectors:
\[
\cos(\vec{a}, \vec{b}) = \frac{\langle \vec{a},  \vec{a}\rangle}{|\vec{a}| |\vec{b}|}
\]
Now we define scalar product of two random vectors as covariation between them:
\[
\langle X, Y \rangle = \Cov(X, Y)
\]
The main characteristics of a random vector are its length and direction.
So, we introduce the length
\[
\sqrt{\Cov{X,X}} = \sqrt{\Var(X)} = \sigma_X
\]
and the angle between two random vectors
\[
\cos(X,Y) = \frac{\Cov(X,Y)}{\sqrt{\Var(X)\Var(Y)}} = \Corr(X,Y)
\]
Note that from the definition of the angle it follows that correlation can range from $-1$ to $1$.

\begin{figure}[h!]
\begin{center}
\includegraphics{images/corr_def.pdf}
\caption{Random vector $X$ of length $\sigma_X$ and random vector $Y$ of length $\sigma_Y$, $\cos \varphi$ is the angle between $X$ and $Y$.}
\label{fig:corr_def}
\end{center}
\end{figure}

Another important geometrical tool is projection. Recall that for any two vectors the scalar product $\langle \vec{a}, \vec{b} \rangle$ can be interpreted as the length of projected $\vec{b}$ multuplied by the length of $\vec{a}$.
The projection itself is $cos(\vec{a}, \vec{b}) \vec{b}$.
Same holds for random vectors. The projection of a random vector $Y$ onto $\{cX| c \in \mathbb{R}\}$ is $\hat Y = \Corr(X,Y) \cdot Y$.

\begin{figure}[h!]
\begin{center}
\includegraphics{images/corr_proj.pdf}
\caption{Random vector $Y$ projected onto random vector $X$.}
\label{fig:corr_proj}
\end{center}
\end{figure}

Looking at Figure~\ref{fig:corr_proj}, we can interpret the square of correlation coefficient.
Using the fact that $cos^2 \varphi$ is the squared ratio of the leg adjacent to $\varphi$ to hypotenuse, we can write
\[
\Corr^2(X,Y) = \frac{\Var(\hat Y)}{\Var(Y)}
\]
as the variance of a random vecor is associated with the square of its length.
Thus, the correlation coefficient squared shows the fraction of variance in $Y$ which can be explained
with the most similar random variable proportional to $X$.

@@

Now consider the case when $\hat \beta_2 < 0$.
Note that the sign of $\beta_1$ does not influence the correlation coefficient sign.
The only difference is that now $\hat y$ is projected onto the span of  $x - \bar x \cdot \mathbf{1}$ not on this vector itself while the projections of $x$ and $y$ remain the same.
Looking at Figure~\ref{fig:corr_negative} we deduce that the angle betwween $y$ and $\hat y$ is compelement to the angle between $x$ and $y$.
Using trigonometric properties, we simplify $\cos(180^{\circ} - \varphi) = -\cos\varphi$ which in turn implies $\sCorr(x,y) = -\sCorr(y,\hat y)$.

@@

\begin{theorem}
A linear regression model with $n$ observations and $k$ explanatory variables including a constant unit vector
\[
y = X \beta + \varepsilon
\]
has the following property
\[
RSS + ESS = TSS
\]
where $RSS = \lVert y - \hat y \rVert^2_2$, $ESS = \lVert \hat y - \bar y \rVert^2_2$, $TSS = \lVert y - \bar y \rVert$.
\end{theorem}

@@

So as to visualize $ESS$ and $TSS$ we first need to visualize vector of averages $\bar y$.
Geometrically this means projecting a vector onto a line spanned by vector $\mathbf{1}$.
It can be shown that the matrix corresponding to projecting onto the line spanned by a vector of all ones has the following form
\[
\frac{\mathbf{1}^T \mathbf{1}}{\mathbf{1} \mathbf{1}^T} = \frac{\begin{pmatrix} 1 & \ldots & 1 \end{pmatrix} \begin{pmatrix} 1 \\ \ldots \\ 1 \end{pmatrix}}{\sum_{i=1}^n 1} = \begin{pmatrix} \frac{1}{n} & \ldots & \frac{1}{n} \\ \vdots & \ddots & \vdots \\ \frac{1}{n} & \ldots & \frac{1}{n} \end{pmatrix}
\]

@@

The final step is to put everything together.
Note that since $y - \hat y$ is perpendicular to $lin(x, \mathbf{1})$ it is also perpendicular to $\hat y - \bar y$ and $\mathbf{1}$ as these vectoros are in $lin(x, \mathbf{1})$.
Then, applying the theorem of three perpendiculars we conclude that vecotr foot of vector $y - \bar y$ is the same point as the foot of the vector $\hat y - \bar y$.
Thus, we obtain a right angle triangle and can apply the Pythagorean theorem for the catheti $\sqrt{RSS}$ and $\sqrt{ESS}$ and the hypothenuse $\sqrt{TSS}$:
\[
\sqrt{RSS} + \sqrt{ESS} = \sqrt{TSS}
\]

\begin{figure}[h!]
\center{\includegraphics{images/tss_sum_final.pdf}}
\caption{$\sqrt{RSS} + \sqrt{ESS} = \sqrt{TSS}$}
\label{fig:tss_final}
\end{figure}

@@

@@

\marginnote{
If the regression contains the intercept, the following equation holds:
\begin{align*}
\hat y  &= X \hat \beta = X (X^T X)^{-1} X^T y \\
&= X (X^T X)^{-1} X^T X \beta + X (X^T X)^{-1} X^T \varepsilon
\end{align*}
Premultiplying both sides by $X^T$, we obtain:
\begin{align*}
X^T \hat y &=  X^T X (X^T X)^{-1} X^T X \beta \\
&+ X^T X (X^T X)^{-1} X^T \varepsilon \\
&= X^T X \beta + X^T \varepsilon
\end{align*}
This is a system of equations. The first row of $X^T$ is $\mathbf{1}$ vector, so we can write out the first equation:
\[
\sum_{i=1}^n \hat y_i = \sum_{i=1}^{n} \sum_{j=1}^{k} x_{ij} \beta_{j}
\]
From the first equation in the system
\[
X^T \hat y = X^T y
\]
we obtain
\[
\sum_{i=1}^{n} \hat y_i = \sum_{i=1}^{n} y
\]
And this finishes the proof:
\[
\frac{1}{n} \sum_{i=1}^{n} y = \frac{1}{n} \sum_{i=1}^{n} \sum_{j=1}^{k} x_{ij} \beta_{j}
\]
}

>>>>

\subsection{Sample correlation}

\marginnote{
\begin{align*}
\sCorr(x,y) &= \frac{\sCov(x,y)}{\sqrt{\sVar(x)\sVar(y)}} \\
&= \frac{\frac{1}{n-1}\sum_{i=1}^n (x_i - \bar x)(y_i - \bar y)}{\sqrt{\frac{1}{n-1} \sum_{i=1}^n (y_i - \bar y)\frac{1}{n-1} \sum_{i=1}^n (\hat y_i - \bar{\hat y})}}
\end{align*}
}

The most crucial part in defining correlation geometrically is definig the dot product as it enables to compute the length of a vecotr:
\[
|\vec{a}| = \sqrt{\langle \vec{a},  \vec{a}\rangle}
\]
and the angle between any two vectors:
\[
\cos(\vec{a}, \vec{b}) = \frac{\langle \vec{a},  \vec{a}\rangle}{|\vec{a}| |\vec{b}|}
\]
Now we define scalar product of two vectors
$x =
\begin{pmatrix}
x_1 \\
\vdots \\
x_n
\end{pmatrix}$
and
$y =
\begin{pmatrix}
y_1 \\
\vdots \\
y_n
\end{pmatrix}$
as a sample covariation between them:
\[
\langle x, y \rangle = \sCov(x, y)
\]
The main characteristics of a vector are its length and direction.
So, we introduce the length
\[
\sqrt{\sCov(x,x)} = \sqrt{\sVar(x)} = \sigma_x
\]
and the angle between two random vectors
\[
\cos(x,y) = \frac{\sCov(x,y)}{\sqrt{\sVar(x)\sVar(y)}} = \sCorr(x,y)
\]
Note that from the definition of the angle it follows that the sample correlation can range from $-1$ to $1$.

\begin{figure}[h!]
\begin{center}
\includegraphics{images/corr_def.pdf}
\caption{Vector $x$ of length $\sigma_x$ and random vector $y$ of length $\sigma_y$, $\cos \varphi$ is the angle between $x$ and $y$.}
\label{fig:corr_def}
\end{center}
\end{figure}

Another important geometrical tool is projection. Recall that for any two vectors the scalar product $\langle \vec{a}, \vec{b} \rangle$ can be interpreted as the length of projected $\vec{b}$ multuplied by the length of $\vec{a}$.
The projection itself is $cos(\vec{a}, \vec{b}) \vec{b}$.
Same holds for the vectors sampled from some distribution.
The projection of such a vector $y$ onto $\{cx| c \in \mathbb{R}\}$ is $\hat y = \sCorr(x,y) \cdot y$.

\begin{figure}[h!]
\begin{center}
\includegraphics{images/corr_proj.pdf}
\caption{Vector $y$ projected onto vector $x$.}
\label{fig:corr_proj}
\end{center}
\end{figure}

Looking at Figure~\ref{fig:corr_proj}, we can interpret the square of sample correlation coefficient.
Using the fact that $cos^2 \varphi$ is the squared ratio of the leg adjacent to $\varphi$ to hypotenuse, we can conclude that
\[
\sCorr^2(x,y) = \frac{\sVar(\hat y)}{\sVar(y)}
\]
as the variance of a vector is associated with the square of its length.
Thus, the sample correlation coefficient squared shows the fraction of variance in $y$ which can be explained
with the most similar vector proportional to $x$.

\subsection{Sample correlation when a constatnt vector added}

\marginnote{
\begin{align*}
\sCorr(x + \alpha \mathbf{1}, y) &= \frac{\sCov(x + \alpha \mathbf{1}, y)}{\sVar(x + \alpha \mathbf{1}) \sVar(y)} \\
&= \frac{\sCov(x,y)\sCov(\alpha \mathbf{1},y)}{\sVar(x)\sVar(y)} \\
&= \frac{\sCov(x,y)}{\sVar(x)\sVar(y)} \\
&= \sCorr(x,y)
\end{align*}
}

\begin{theorem}
Adding a vector of constants does not affect the sample correlation coefficient:
\[
\sCorr(x + \alpha \mathbf{1}, y) = \sCorr(x,y)
\]
where $\alpha \in \mathbb{R}$.
\end{theorem}

\begin{proof}
Firstly, we project vectors $x$ and $y$ onto $Lin^{\perp}(\mathbf{1})$ in order to get $x^c = x - \bar x$ and $y^c = y - \bar y$
(`c' stands for `centred'). It can be shown that the matrix corresponding to projecting onto the line spanned by a vector of all ones has the following form
\[
\frac{\mathbf{1}^T \mathbf{1}}{\mathbf{1} \mathbf{1}^T} = \frac{\begin{pmatrix} 1 & \ldots & 1 \end{pmatrix} \begin{pmatrix} 1 \\ \ldots \\ 1 \end{pmatrix}}{\sum_{i=1}^n 1} = \begin{pmatrix} \frac{1}{n} & \ldots & \frac{1}{n} \\ \vdots & \ddots & \vdots \\ \frac{1}{n} & \ldots & \frac{1}{n} \end{pmatrix}
\]
Thus, projecting onto an orthogonal subspace is equivalent to substracting the projected vector, i.e., the vector of averages, from the original one.

Also note that the angle $\varphi$ between original and centred vectors remains the same.
The results of this step is shown in Figure~\ref{fig:corr_xyc}.

\begin{figure}
  \centering
  \includegraphics{images/corr_const_cent.pdf}
  \caption{Centred vectors $x^c$ and $y^c$}
  \label{fig:corr_xyc}
\end{figure}

Then we need to derive a new vector $\tilde x$ with constants added to each component.
Geometrically adding a vector of costants means adding a vector of all ones scaled by $\alpha \in \mathbb{R}$, i.e., $\alpha \mathbf{1}$.
Then the new vector $\tilde x$ can be broken up into a sum of $\alpha \mathbf{1}$ and $\beta x$, $\alpha, \beta \in \mathbb{R}$, which can be seen in Figure~\ref{fig:corr_tildex}.
After that we will project this new vector $\tilde x$ onto $Lin^{\perp}(\mathbf{1})$.
By the properties of projection it is of no difference whether to project the whole vector $\tilde x$ or project its parts $\alpha \mathbf{1}$ and $\beta x$ — the result is the same.
So, while $\beta x$ is projected onto a span of $x^c$, the projection of $\alpha \mathbf{1}$ onto the orthohgonal space $Lin^{\perp}(\mathbf{1})$ yields zero as demonstrated in Figure~\ref{fig:corr_tildex}.
Moreover, it follows that the angle between $\tilde x$ and $y$ is still $\varphi$.

\begin{figure}
  \centering
  \includegraphics{images/corr_const_br_up.pdf}
  \caption{Decomposition and projection of $\tilde x$}
  \label{fig:corr_tildex}
\end{figure}

Finally, putting everything together we finish the proof:
\[
\sCorr(x + \alpha \mathbf{1}, y) = \sCorr(x,y)
\]

\begin{figure}
  \centering
  \includegraphics{images/corr_const_final.pdf}
  \caption{$\sCorr(x + \alpha \mathbf{1}, y) = \sCorr(x,y)$ as the corresponding angles are equal.}
  \label{fig:corr_final}
\end{figure}

@@

Now consider the case when $\hat \beta_2 < 0$.
Note that the sign of $\beta_1$ does not influence the correlation coefficient sign.
The only difference is that now $\hat y$ is projected onto the span of  $x - \bar x \cdot \mathbf{1}$ and not on this vector itself while the projections of $x$ and $y$ remain the same.
Looking at Figure~\ref{fig:corr_negative} we deduce that the angle betwween $y$ and $\hat y$ is compelement to the angle between $x$ and $y$.
Using trigonometric properties, we simplify $\cos(180^{\circ} - \varphi) = -\cos\varphi$ which in turn implies $\sCorr(x,y) = -\sCorr(y,\hat y)$.

@@

\begin{theorem}
A linear regression model with $n$ observations and $k$ explanatory variables including a constant unit vector
\[
y = X \beta + \varepsilon
\]
has the following property
\[
RSS + ESS = TSS
\]
where $RSS = \lVert y - \hat y \rVert^2_2$, $ESS = \lVert \hat y - \bar y \rVert^2_2$, $TSS = \lVert y - \bar y \rVert^2_2$.
\end{theorem}

@@

So as to visualize $ESS$ and $TSS$ we first need to visualize vector of averages $\bar y$.
Geometrically this means projecting a vector onto a line spanned by vector $\mathbf{1}$.

@@

The final step is to put everything together.
Note that since $y - \hat y$ is perpendicular to $Lin(x, \mathbf{1})$ it is also perpendicular to $\hat y - \bar y$ and $\mathbf{1}$ as these vectoros are in $Lin(x, \mathbf{1})$.
Then, applying the theorem of three perpendiculars we conclude that the foot of vector $y - \bar y$ is the same point as the foot of the vector $\hat y - \bar y$.
Thus, we obtain a right angle triangle and can apply the Pythagorean theorem for the catheti $\sqrt{RSS}$ and $\sqrt{ESS}$ and the hypothenuse $\sqrt{TSS}$:
\[
\sqrt{RSS} + \sqrt{ESS} = \sqrt{TSS}
\]

\begin{figure}[h!]
\center{\includegraphics{images/tss_sum_final.pdf}}
\caption{$(\sqrt{RSS})^2 + (\sqrt{ESS})^2 = (\sqrt{TSS})^2$}
\label{fig:tss_final}
\end{figure}

@@

\marginnote{
\begin{align*}
\sCorr^2(y,\hat y) &= \left(\frac{\sCov(y, \hat y)}{\sqrt{\sVar(y)\sVar(\hat y)}}\right)^2 \\
&= \frac{\sCov(y, \hat y) \sCov(y, \hat y)}{\sVar(y)\sVar(\hat y)} \\
&= \frac{\sCov(\hat y + e, \hat y) \sCov(\hat y + e, \hat y)}{\sVar(y)\sVar(\hat y)} \\
&= \frac{(\sCov(\hat y, \hat y) + \sCov(e, \hat y))(\sCov(\hat y, \hat y) + \sCov(e, \hat y))}{\sVar(y)\sVar(\hat y)} \\
&= \frac{\sVar(\hat y) \sVar(\hat y)}{\sVar(y)\sVar(\hat y)} \\
&= \frac{\sVar(\hat y)}{\sVar(y)} \\
&= \frac{ESS}{TSS} \\
&= R^2
\end{align*}
}

\begin{theorem}
A linear regression model with $n$ observations and $k$ explanatory variables including a constant unit vector
\[
y = X \beta + \varepsilon
\]
has the following property
\[
R^2 = \sCorr^2(y, \hat y)
\]
\end{theorem}

\begin{proof}
Proving this theorem geometrically means showing that the determination coefficient can be interpreted as some squared angle
which happens to be eqaul to the squared angle betwen $y$ and $\hat y$.

Consider Figure~\ref{fig:tss_final} from the previous proof.
It was shown there that the vectors $y - \bar y$, $y - \hat y$ and $\hat y - \bar y$ form a right triangle.
Having defined the determination coefficient as
\[
R^2 = \frac{ESS}{TSS}
\]
we conclude that its geometric interpretaion is
\[
\sqrt{R^2} = \frac{\sqrt{ESS}}{\sqrt{TSS}} = \cos \varphi
\]
as shown in Figure~\ref{fig:r_sq_angle}.

\begin{figure}
  \center{\includegraphics{images/r_sq_ss.pdf}}
  \caption{Determination coefficient as squared $\cos \varphi$}
  \label{fig:r_sq_angle}
\end{figure}

Recall that the sample correlation coefficient two vectors was defined earlier as the angle between these two vectors.
Thus, we conclude that $\sCorr(y, \hat y)$ is the angle between $y$ and $\hat y$ which is also eqaul to $\cos \varphi$.
Finally, squaring both sides, we obtain
\[
R^2 = \sCorr^2(y, \hat y)
\]
\end{proof}

@@

\marginnote{
If the regression contains the intercept, the following equation holds:
\begin{align*}
\hat y  &= X \hat \beta \\
&= X (X^T X)^{-1} X^T y \\
&= X (X^T X)^{-1} X^T X \beta + X (X^T X)^{-1} X^T \varepsilon
\end{align*}
Premultiplying both sides by $X^T$, we obtain:
\begin{align*}
X^T \hat y &=  X^T X (X^T X)^{-1} X^T X \beta \\
&+ X^T X (X^T X)^{-1} X^T \varepsilon \\
&= X^T X \beta + X^T \varepsilon
\end{align*}
This is a system of equations. The first row of $X^T$ is $\mathbf{1}$ vector, so we can write out the first equation:
\[
\sum_{i=1}^n \hat y_i = \sum_{i=1}^{n} \sum_{j=1}^{k} x_{ij} \beta_{j}
\]
From the first equation in the system
\[
X^T \hat y = X^T y
\]
we obtain
\[
\sum_{i=1}^{n} \hat y_i = \sum_{i=1}^{n} y
\]
And this finishes the proof:
\[
\frac{1}{n} \sum_{i=1}^{n} y = \frac{1}{n} \sum_{i=1}^{n} \sum_{j=1}^{k} x_{ij} \beta_{j}
\]
}
